---
title: "560 Lecture 14 - Maximum a Posteriori Estimation"
author: "Su-In Lee"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal of this exercise is to understand the impact of hyperparameters in the Bernoulli experiment we discussed in class.  We will compare the likelihood P(D|p) with the posterior P(p|D) for varying alpha and beta values. Since P(D) does not depend on p, we will ignore this term in the posterior P(p|D).

#### Hint: here is how to generate a function in R:
myfunction <- function(arg1, arg2, ...){
  statments
  return(object)
}

#### Hint: gamma function in R
gamma(x)

#### Let's first specify the values of nh and nt. Later, we will adjust these values and see how the mode of the likelihood function changes.

```{r}
nh <- 100
nt <- 50
```

#### Let's specificy the hyperparameters (pseudo counts)
```{r}
alpha <- 1000
beta <- 1000
```

#### Let's now define the functions for the likelihood and posteria.
```{r}
likelihood <- function(p, nh, nt){
  return()
}

posterior <- function(p, nh, nt, alpha, beta){
  return()
}
```

#### Let's plot the log-likelihood function as a function of p for varing values of hyperparameters. For example, try the following as well.

```{r}
nh <- 100
nt <- 50
alpha <- 30
beta <- 30
```

#### Describe the difference.
