---
title: "560 Lecture 18 - One Way ANOVA"
author: "Su-In Lee"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We are going to do two one-way analysis of variance (ANOVA). We use the "lm" 
command which fits a "linear model" and then the "anova" command which presents 
the analysis of variance table.

In each case see if you can figure out what the ANOVA table means, 
what tests what, and why the number of degrees of freedom is what it is.


#### Case 1.

I have made a data frame called "example_ANOVA.txt" which you will find
on the web site.  (It is artificial data).  Download it, save it to
the folder (directory) in which you will be working for this exercise.

It has three columns.  The first is the observation number.
The second is the group number (grp), the third is the observation (x).
These variable names are known to the program after the data frame is read in.

Do a one-way ANOVA.  You need to read the table using
the usual command

```{r}
ff <- read.table(file="http://www.cs.washington.edu/homes/suinlee/genome560/data/example_ANOVA.txt")      
```

and then use "lm" and "anova" commands:

```{r}
mm <- lm(x ~ factor(grp), data=ff)     # checks effect of variable "grp"
# which is a "factor" (not a number)
anova(mm)                        # reports the ANOVA table
# Try to figure out the ANOVA table.
mm                               # If you just print "mm" what does it show?
```

#### Case 2.

These are (real) measurements of adductor muscle scar
length on shells of mussels collected at a number of localities.
A one-way ANOVA tests whether those localities differ.
The numerical value of variable V1 is to be used as a "factor",
so that it simply indicates group membership and is not
a numerical value of interest.

Read the data frame (and also be sure to take a look at it):

```{r}
gg <- read.table(file="http://www.cs.washington.edu/homes/suinlee/genome560/data/mussels.txt")      

m2 <- lm(V2 ~ V1, data=gg)           # fit the linear model
# It knows to use V1 as a "factor"
# but you could use  V2 ~ factor(V1)  too
anova(m2)                            # print the ANOVA table
m2                                   # also the estimated effects
```

Do the localities differ?  What accounts for the degrees of freedom?
Do we need to have equal numbers of data points in each group?

