---
title: "560 Lecture 8 - Nonparametric Location Tests"
author: "Douglas M. Fowler"
date: "3/26/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
```

## Sign Test

#### First let's install and load the appropriate package
```{r}
#install.packages("BSDA")
library(BSDA)
```

#### Now, let's generate some data to test
```{r}
x = rnorm(100, 2)
```

#### Let's test to see if this data really does come from a distribution with a median of 2
```{r}
SIGN.test(x, md = 2)
```

## Wilcoxon Signed Rank Test

#### Now let's try the Wilcoxon Signed Rank Test on our data
```{r}
wilcox.test(x, mu = 2)
```

#### Interesting - using the rank data in addition to the sign data gave us much better precision.

#### Now, you try with paired data.  First, generate some random data from the normal distribution, calculate the difference and do the appropriate test.
```{r}

```

## Wilcoxon Rank Sum Test

#### Now, assume that your made up normally distribution data is actually a set that you'd like to compare to the first data set x.  Use wilcox.test to test whether they are the same.  
```{r}

```

####  Now make random normal data set of 1000 elements with a mean of 2 and a random gamma data set whose shape parameter is 2 (will also have an expected value of 2).  Make histograms of each, marking the mean and median.  
```{r}

```

#### What do you think will happen when you do a Wilcoxon rank sum test with these data?  Give it a try and see.
```{r}

```

#### Next, do a WRST on the WW domain reported_effect scores vs the Pab1 reported_effect scores
```{r}
ww_data = read.table(file="http://faculty.washington.edu/dfowler/teaching/2017/GNOM560/560_ww_data.txt", header = T, sep = '\t')
pab_data = read.table(file="http://faculty.washington.edu/dfowler/teaching/2017/GNOM560/560_pab_data.txt", header = T, sep = '\t')


```

#### Do you think it's a good idea to use this test for these data?  If so, why?  If not, why not?


## Kruskal-Wallis Test

#### OK, now let's give the Kruskal-Wallis test a try.  Create three random gamma-distributed data set with 100 elements and an identical shape parameter of your choice
```{r}

```

#### Now, use kruskal.test() to verify that the medians are the same 
```{r}

```

#### OK, now double one of the location parameters and test again
```{r}

```

#### Use pairwise.wilcox.test() to see which of our three data sets is different from the others (note the automatic correction for multiple hypothesis testing)
```{r}

```

#### To get a sense of the power of the KW test, try varying the shape parameter in increments of 1% up or down and find the threshold for detecting the difference
```{r}

```

# This is interesting... to draw a conclusion try repeating this exercise 1000 times
```{r}

```