---
title: "560 HW #1"
author: "David Bacsik"
date: "April 10, 2018"

output: html_document
---
### *due date: Thursday, April 12, 2018*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part 1
#### 1.1a
Load the master data file and select the data I want from it. I am going to use the ww_data we used in the lab.

```{r}
master_data = read.table(file="560_dataset_reduced.txt", header = T, sep = '\t')

head(master_data)

ww_data = master_data[which(master_data$dms_id == 'WW_domain'),]

head(ww_data)
```

#### 1.1b
Now, I want to calculate some summary statistics about the functional scores of each variant.

```{r}
func_scores = ww_data$reported_effect


fs.range = range(func_scores)

fs.mean = mean(func_scores)

fs.var = var(func_scores)

fs.sd = sd(func_scores)

quant_list = c(0.025, 0.05, 0.25, 0.5, 0.75, 0.95, 0.975)
fs.quants = quantile(func_scores, probs=quant_list)
```

The range of functional scores is: `r fs.range`

The mean of functional scores is: `r fs.mean`

The variance of functional scores is: `r fs.var`

The standard deviation of functional scores is: `r fs.sd`

Quantiles:

```{r} 
print(fs.quants)
```

Are the 25% and 75% quantiles approximately .67 sd from the mean?'

```{r}
approx_25 = mean(func_scores)-0.67448*sd(func_scores)
approx_75 = mean(func_scores)+0.67488*sd(func_scores)
```

In a normal distribution, the 25th percentile would be approximately `r approx_25`, and in our data it is approximately `r quantile(func_scores, probs=c(0.25))`. The 75th percentile woudl be approximately `r approx_75`, and in our data it is approximately `r quantile(func_scores, probs=c(0.75))`.

These numbers are pretty close, so based on this, I conclude that our data are approximately normally distributed.

#### 1.1c
I want to visualize the distribution of our data. To do this, I will make a histogram of the functional scores with 50 bins.

```{r}

hist(func_scores, 50, freq=TRUE, main='WW DMS Functional Scores', xlab='Score', ylab='Frequency')
```

#### 1.1d
I'd like to see if there are broad trends to mutation to a specific amino acid. For each of the 20 amino acids, I will calculate the mean effect of mutation to that amino acid.

```{r}

aa_list = c()