---
title: "week1_lab2"
author: "db"
date: "March 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 1, Lab 2

This is the second lab of week 1. Focused on R Markdown files.

```{r}
ww_data = read.table(file="560_ww_data.txt", header = T, sep = '\t')
```

Get some summary info about this dataset. str: (all variables, number of observations)
```{r}
str(ww_data)
```

Get some summary info about this dataset. head: (first 10 rows or so)
```{r}
head(ww_data)
```

Get some summary info about this dataset. names: (all column names)
```{r}
names(ww_data)
```

## Descriptive statistics
Now, let's get some summary stats on this data.
For a broad overview of expression for each feature in the data set we can use the summary() function

```{r}
summary(ww_data)
```

We can also have a look at the first row of the master data frame (e.g. the first mutation):
```{r}
ww_data[1,]
```

Let's extract the data for the functional scores.

Subsetting and data types can be tricky in R. Here, we want to use the numeric values for calculations, so we use as.vector() with the parameter mode=“numeric” to get a vector of numbers.

Then, use is to get the variable type and verify we've created what we expected.

```{r}
ps1 = as.vector(ww_data$reported_effect, mode='numeric')

is(ps1)
```

Now, let's do some calculations on the functional score.
First, let's find the average, range, and standard deviation.
```{r}
ps1.mean = mean(ps1)
ps1.range = range(ps1)
ps1.sd = sd(ps1)

ps1.mean
ps1.range
ps1.sd
```

Also, we can subset the data on conditionals for filtering.
Let's collect the functional effect for for just mutations to alanine or just tryptophan, and calculate some summary stats.

```{r}
just_ala = ww_data[which(ww_data$aa2 == 'A'),]$reported_effect
mean(just_ala)

just_trp = ww_data[which(ww_data$aa2 == 'W'),]$reported_effect
mean(just_trp)
```

